---
title: "Welton_4_state_forward"
format: html
---

```{r}
### 1. Load packages
library(R2jags)
library(tidyverse)
library(coda)
library(abind)


model_code <- textConnection("
model{
#Multinomial likelihood for observed data
for (i in 1:3){ r[i,1:4] ~ dmulti(P[i,1:4],n[i]) }
#Find transition probabilities (for given time) in terms of rates.
Sum transition rate from #state i, lambda[i] denoted s[i] for
brevity.
for (i in 1:3){
s[i]<- sum(G[i,(i+1):4])
e[i]<- exp(-s[i]*t.obs)
P[i,4]<- 1 - sum(P[i,1:3])
#Give exponential priors for unknown transition rate parameters
for (j in (i+1):4){G[i,j] ~ dexp(.001)}
G[i,i]<- -lambda[i]
}
P[1,1]<-e[1]
P[1,2]<- G[1,2]*(e[2] - e[1])/(s[1] - s[2])
P[1,3]<- G[1,3]*(e[3] - e[1])/(s[1] - s[3]) + G[1,2]*G[2,3]*((s[1]-s[2])*e[3]
- (s[1]-s[3])*e[2] + (s[2]-s[3])*e[1])/((s[1]-s[2])*(s[1]-s[3])*(s[2]-s[3]))
P[2,2]<-e[2]
P[2,3]<- G[2,3]*(e[3] - e[2])/(s[2] - s[3])
P[3,3]<-e[3]
P[2,1]<-0
P[3,1]<-0
P[3,2]<-0

#Find P(t) where t = t.new
for (i in 1:3){
e.t[i]<- exp(-s[i]*t.new)
P.t[i,4]<- 1 - sum(P.t[i,i:3])
}
P.t[1,1]<-e.t[1]
P.t[1,2]<- G[1,2]*(e.t[2] - e.t[1])/(s[1] - s[2])
P.t[1,3]<- G[1,3]*(e.t[3] - e.t[1])/(s[1] - s[3])
+ G[1,2]*G[2,3]*((s[1]-s[2])*e.t[3] - (s[1]-s[3])*e.t[2]
+ (s[2]-s[3])*e.t[1])/((s[1]-s[2])*(s[1]-s[3])*(s[2]-s[3]))
P.t[2,2]<-e.t[2]
P.t[2,3]<- G[2,3]*(e.t[3] - e.t[2])/(s[2] - s[3])
P.t[3,3]<-e.t[3]
}
")

#Data
list( r=structure(.Data=c(20,6,4,2,
0,5,5,3,
0,0,3,7), .Dim=c(3,4)),
n=c(32,13,10), t.obs=2, t.new=.25)
#Initial Values
list(G=structure(.Data=c(NA,1,1,1,
NA,NA,1,1,
NA,NA,NA,1
), .Dim=c(3,4)))

### 4. Fit model
set.seed(42)
results <- jags(
  data = data_list,
  inits = inits,
  parameters.to.save = parameters.to.save,
  model.file = model_code,
  n.chains = 1,
  n.iter = 10000,
  n.burnin = 1000,
  n.thin = 10
)
```




NEW

```{r}
library(tibble)
library(ggplot2)

```

```{r}
library(R2jags)
library(tidyverse)
library(coda)
library(abind)

# JAGS model code
model_code <- textConnection("
model {
  for (i in 1:3) {
    r[i, 1:4] ~ dmulti(P[i, 1:4], n[i])
  }

  for (i in 1:3) {
    for (j in 1:4) {
      is_offdiag[i, j] <- step(j - i)
      G_temp[i, j] ~ dunif(0, 5)
      G[i, j] <- ifelse(j == i, -sum(G_temp[i, (i+1):4]),
                     ifelse(j > i, G_temp[i, j], 0))
    }

    s[i] <- -G[i, i]
    e[i] <- exp(-s[i] * t_obs)
  }

  # Transition probabilities at t_obs
  P[1,1] <- e[1]
  P[1,2] <- G[1,2] * (e[2] - e[1]) / (s[1] - s[2])
  P[1,3] <- G[1,3] * (e[3] - e[1]) / (s[1] - s[3]) +
            G[1,2] * G[2,3] * ((s[1] - s[2]) * e[3] -
                               (s[1] - s[3]) * e[2] +
                               (s[2] - s[3]) * e[1]) /
            ((s[1] - s[2]) * (s[1] - s[3]) * (s[2] - s[3]))
  P[1,4] <- 1 - sum(P[1, 1:3])

  P[2,1] <- 0
  P[2,2] <- e[2]
  P[2,3] <- G[2,3] * (e[3] - e[2]) / (s[2] - s[3])
  P[2,4] <- 1 - sum(P[2, 1:3])

  P[3,1] <- 0
  P[3,2] <- 0
  P[3,3] <- e[3]
  P[3,4] <- 1 - sum(P[3, 1:3])

  # Transition probabilities at t_new
  for (i in 1:3) {
    e_t[i] <- exp(-s[i] * t_new)
  }

  P_t[1,1] <- e_t[1]
  P_t[1,2] <- G[1,2] * (e_t[2] - e_t[1]) / (s[1] - s[2])
  P_t[1,3] <- G[1,3] * (e_t[3] - e_t[1]) / (s[1] - s[3]) +
              G[1,2] * G[2,3] * ((s[1] - s[2]) * e_t[3] -
                                 (s[1] - s[3]) * e_t[2] +
                                 (s[2] - s[3]) * e_t[1]) /
              ((s[1] - s[2]) * (s[1] - s[3]) * (s[2] - s[3]))
  P_t[1,4] <- 1 - sum(P_t[1, 1:3])

  P_t[2,1] <- 0
  P_t[2,2] <- e_t[2]
  P_t[2,3] <- G[2,3] * (e_t[3] - e_t[2]) / (s[2] - s[3])
  P_t[2,4] <- 1 - sum(P_t[2, 1:3])

  P_t[3,1] <- 0
  P_t[3,2] <- 0
  P_t[3,3] <- e_t[3]
  P_t[3,4] <- 1 - sum(P_t[3, 1:3])
}
")

# Data
data_list <- list(
  r = matrix(c(
    20, 6, 4, 2,
    0, 5, 5, 3,
    0, 0, 3, 7
  ), nrow = 3, byrow = TRUE),
  n = c(32, 13, 10),
  t_obs = 2,
  t_new = 0.25
)

# Initial values
inits <- function() {
  list(G_temp = matrix(runif(12, 0.1, 1), nrow = 3, ncol = 4))
}

# Parameters to monitor
parameters.to.save <- c("G", "P", "P_t")

# Run JAGS
set.seed(42)
results <- jags(
  data = data_list,
  inits = inits,
  parameters.to.save = parameters.to.save,
  model.file = model_code,
  n.chains = 1,
  n.iter = 10000,
  n.burnin = 1000,
  n.thin = 10
)

# Summary
print(results)

```


Nice—let’s visualize those posteriors! To plot the posterior distributions of the transition probabilities from your JAGS model, we’ll use the coda output stored in results$BUGSoutput$sims.list. Here’s how to extract and plot the marginal posterior densities for selected P[i,j] and P_t[i,j] values:

```{r}
# Extract posterior samples
sims <- results$BUGSoutput$sims.list

# Choose a few transition probabilities to illustrate
library(ggplot2)
library(tidyr)
library(dplyr)

# Convert to tidy format for plotting
plot_data <- tibble(
  `P[1,2]` = sims$P[,1,2],
  `P[1,3]` = sims$P[,1,3],
  `P[2,3]` = sims$P[,2,3],
  `P_t[1,3]` = sims$P_t[,1,3],
  `P_t[2,3]` = sims$P_t[,2,3]
) %>%
  pivot_longer(everything(), names_to = "Parameter", values_to = "Value")

# Plot posterior densities
ggplot(plot_data, aes(x = Value, fill = Parameter)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ Parameter, scales = "free") +
  labs(title = "Posterior Distributions of Transition Probabilities",
       x = "Probability", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

```

Perfect—let’s bring those survival curves to life! We'll compute survival probabilities starting from each transient state (states 1, 2, and 3) over a grid of time points. To do that, we'll re-run the model for a vector of time values and extract P_t[i,4], which gives the cumulative probability of absorption (i.e. transitioning to the absorbing state by time t). The survival probability is simply 1 - P_t[i,4].

Here’s how to set this up in R:

```{r}
jags_model_string <- "
model {
  for (i in 1:3) {
    r[i, 1:4] ~ dmulti(P[i, 1:4], n[i])
  }

  for (i in 1:3) {
    for (j in 1:4) {
      is_offdiag[i, j] <- step(j - i)
      G_temp[i, j] ~ dunif(0, 5)
      G[i, j] <- ifelse(j == i, -sum(G_temp[i, (i+1):4]),
                     ifelse(j > i, G_temp[i, j], 0))
    }

    s[i] <- -G[i, i]
    e[i] <- exp(-s[i] * t_obs)
  }

  # Transition matrix at t_obs (needed but not used here)
  P[1,1] <- e[1]
  P[1,2] <- G[1,2]*(e[2] - e[1]) / (s[1] - s[2])
  P[1,3] <- G[1,3]*(e[3] - e[1]) / (s[1] - s[3]) +
            G[1,2]*G[2,3]*((s[1] - s[2])*e[3] -
                           (s[1] - s[3])*e[2] +
                           (s[2] - s[3])*e[1]) /
            ((s[1] - s[2])*(s[1] - s[3])*(s[2] - s[3]))
  P[1,4] <- 1 - sum(P[1, 1:3])
  P[2,1] <- 0
  P[2,2] <- e[2]
  P[2,3] <- G[2,3]*(e[3] - e[2]) / (s[2] - s[3])
  P[2,4] <- 1 - sum(P[2, 1:3])
  P[3,1] <- 0
  P[3,2] <- 0
  P[3,3] <- e[3]
  P[3,4] <- 1 - sum(P[3, 1:3])

  # Transition probabilities at t_new
  for (i in 1:3) {
    e_t[i] <- exp(-s[i] * t_new)
  }

  P_t[1,1] <- e_t[1]
  P_t[1,2] <- G[1,2]*(e_t[2] - e_t[1]) / (s[1] - s[2])
  P_t[1,3] <- G[1,3]*(e_t[3] - e_t[1]) / (s[1] - s[3]) +
              G[1,2]*G[2,3]*((s[1] - s[2])*e_t[3] -
                             (s[1] - s[3])*e_t[2] +
                             (s[2] - s[3])*e_t[1]) /
              ((s[1] - s[2])*(s[1] - s[3])*(s[2] - s[3]))
  P_t[1,4] <- 1 - sum(P_t[1, 1:3])

  P_t[2,1] <- 0
  P_t[2,2] <- e_t[2]
  P_t[2,3] <- G[2,3]*(e_t[3] - e_t[2]) / (s[2] - s[3])
  P_t[2,4] <- 1 - sum(P_t[2, 1:3])

  P_t[3,1] <- 0
  P_t[3,2] <- 0
  P_t[3,3] <- e_t[3]
  P_t[3,4] <- 1 - sum(P_t[3, 1:3])
}
"
writeLines(jags_model_string, "markov_model.bug")

```

```{r}
time_grid <- seq(0.1, 5, by = 0.1)
surv_curves <- matrix(NA, nrow = length(time_grid), ncol = 3)
colnames(surv_curves) <- paste0("State ", 1:3)

data_list <- list(
  r = matrix(c(20, 6, 4, 2, 0, 5, 5, 3, 0, 0, 3, 7), nrow = 3, byrow = TRUE),
  n = c(32, 13, 10),
  t_obs = 2,
  t_new = NA  # We'll overwrite this in each iteration
)

inits <- function() {
  list(G_temp = matrix(runif(12, 0.1, 1), nrow = 3, ncol = 4))
  
}

inits_list <- list(inits(), inits(), inits())


for (i in seq_along(time_grid)) {
  data_list$t_new <- time_grid[i]


 results <- jags(
  data = data_list,
  inits = inits_list,
  parameters.to.save = parameters.to.save,
  model.file = "markov_model.bug",
  n.chains = 3,
  n.iter = 10000,
  n.burnin = 1000,
  n.thin = 10
)


  pt_samples <- sim$BUGSoutput$sims.list$P_t
  surv_curves[i, ] <- apply(pt_samples[, , 4], 2, function(x) 1 - mean(x))
}

```


Survival curves

```{r}
# Pre-allocate matrix for survival probabilities
time_grid <- seq(0.1, 5, by = 0.2)
surv_curves <- matrix(NA, nrow = length(time_grid), ncol = 3)
colnames(surv_curves) <- paste0("State ", 1:3)

# Static data that doesn't change
data_static <- list(
  r = matrix(c(20, 6, 4, 2, 0, 5, 5, 3, 0, 0, 3, 7), nrow = 3, byrow = TRUE),
  n = c(32, 13, 10),
  t_obs = 2
)

for (i in seq_along(time_grid)) {
  data_now <- data_static
  data_now$t_new <- time_grid[i]

  sim <- jags(
    data = data_now,
    inits = list(inits(), inits(), inits()),
    parameters.to.save = "P_t",
    model.file = "markov_model.bug",
    n.chains = 3,
    n.iter = 3000,
    n.burnin = 500,
    n.thin = 2,
    quiet = TRUE
  )

  pt <- sim$BUGSoutput$sims.list$P_t
  surv_curves[i, ] <- apply(pt[, , 4], 2, function(x) 1 - mean(x))
}

```


```{r}
library(tibble)
library(tidyr)
library(ggplot2)

surv_df <- as_tibble(surv_curves) %>%
  mutate(Time = time_grid) %>%
  pivot_longer(-Time, names_to = "State", values_to = "Survival")

ggplot(surv_df, aes(x = Time, y = Survival, color = State)) +
  geom_line(size = 1.1) +
  labs(title = "Posterior Mean Survival Curves by Starting State",
       x = "Time", y = "Survival Probability") +
  theme_minimal()
```


Nice—let’s visualize those posteriors! To plot the posterior distributions of the transition probabilities from your JAGS model, we’ll use the coda output stored in results$BUGSoutput$sims.list. Here’s how to extract and plot the marginal posterior densities for selected P[i,j] and P_t[i,j] values:

```{r}
# Extract posterior samples
sims <- results$BUGSoutput$sims.list

# Choose a few transition probabilities to illustrate
library(ggplot2)
library(tidyr)
library(dplyr)

# Convert to tidy format for plotting
plot_data <- tibble(
  `P[1,2]` = sims$P[,1,2],
  `P[1,3]` = sims$P[,1,3],
  `P[2,3]` = sims$P[,2,3],
  `P_t[1,3]` = sims$P_t[,1,3],
  `P_t[2,3]` = sims$P_t[,2,3]
) %>%
  pivot_longer(everything(), names_to = "Parameter", values_to = "Value")

# Plot posterior densities
ggplot(plot_data, aes(x = Value, fill = Parameter)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ Parameter, scales = "free") +
  labs(title = "Posterior Distributions of Transition Probabilities",
       x = "Probability", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

```


DIAGNOSTICS

```{r}
library(coda)
mcmc_samples <- as.mcmc(results)

traceplot(mcmc_samples[, grep("G\\[", varnames(mcmc_samples))], 
          col = "steelblue", 
          main = "Traceplots for G parameters")

traceplot(mcmc_samples[, grep("P_t\\[", varnames(mcmc_samples))], 
          col = "forestgreen", 
          main = "Traceplots for P_t transition probabilities")

autocorr.plot(mcmc_samples[, grep("G\\[", varnames(mcmc_samples))], 
              lag.max = 50, 
              auto.layout = FALSE)

densplot(mcmc_samples[, grep("G\\[", varnames(mcmc_samples))], 
         show.obs = FALSE, 
         col = "darkorange")

gelman.diag(mcmc_samples, multivariate = FALSE)



```

