---
title: "ROC"
author: JB Rickert and my Google AI friend
format: html
---


## Logistic Regression Example 


```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "R Packages"
library(tidyverse)
library(tidymodels)
library(MASS)
```

```{r}
# Load the biopsy dataset
data(biopsy, package = "MASS")

names(biopsy) <- c("ID",
"clump thickness",
"uniformity of cell size",
"uniformity of cell shape",
"marginal adhesion",
"single epithelial cell size",
"bare nuclei",
"bland chromatin",
"normal nucleoli",
"mitoses",
"class")


# Clean column names and convert the outcome to a factor
biopsy <- biopsy %>%
  janitor::clean_names() %>%
  mutate(outcome = factor(class, levels = c("malignant", "benign")))
```


### Fit the Predictive Model

```{r}
# Split the data (a best practice, though not strictly required for this example)
biopsy_split <- initial_split(biopsy, prop = 0.8)
biopsy_train <- training(biopsy_split)
biopsy_test <- testing(biopsy_split)

# Define and fit a logistic regression model
logistic_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(outcome ~ clump_thickness + uniformity_of_cell_size + uniformity_of_cell_shape +
        marginal_adhesion + bare_nuclei, data = biopsy_train)

# Generate predictions on the test set
biopsy_predictions <- predict(logistic_fit, new_data = biopsy_test, type = "prob") %>%
  bind_cols(biopsy_test) %>% drop_na
```

### Manually Create the ROC Curve
```{r}
# Create a sequence of thresholds from 0 to 1
thresholds <- seq(0, 1, by = 0.01)

# Calculate TPR and FPR for each threshold
manual_roc_data <- map_dfr(thresholds, ~ {
  # For a given threshold, classify predictions as "malignant" or "benign"
  predictions_at_threshold <- biopsy_predictions %>%
    mutate(
      .pred_class = factor(
        if_else(.pred_malignant >= .x, "malignant", "benign"),
        levels = c("malignant", "benign")
      )
    )

  # Create a confusion matrix
  conf_mat <- conf_mat(predictions_at_threshold, truth = outcome, estimate = .pred_class)

  # Extract true positives (TP), false positives (FP), etc.
  tp <- conf_mat$table["malignant", "malignant"]
  fp <- conf_mat$table["benign", "malignant"]
  tn <- conf_mat$table["benign", "benign"]
  fn <- conf_mat$table["malignant", "benign"]

  # Calculate TPR and FPR
  tpr <- tp / (tp + fn)
  fpr <- fp / (fp + tn)

  # Return a tibble row for this threshold
  tibble(
    .threshold = .x,
    sensitivity = tpr, # TPR is equivalent to sensitivity
    `1-specificity` = fpr # FPR is equivalent to 1-specificity
  )
})
```

## Density check to see labels are correct

```{r}
biopsy_predictions |> ggplot(aes(x=.pred_malignant, col=outcome)) + geom_density()
```



## Plot Manual ROC Curve
```{r}

# Plot the Manual ROC Curve
ggplot(manual_roc_data, aes(x = `1-specificity`, y = sensitivity)) +
  geom_line(size = 1) +
  geom_abline(linetype = "dashed") +
  scale_x_continuous(limits = c(0,0.2)) +
  labs(
    title = "Manual ROC Curve",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal()

```

In the tidyverse ecosystem, the yardstick package calculates the ROC curve by evaluating all unique probability thresholds present in the data, as well as -infinity and +infinity. This is supposed to provide a more precise curve than manually setting a fixed number of thresholds.



```{r}
# Calculate ROC curve data
roc_data <- biopsy_predictions %>%
  roc_curve(truth = outcome, .pred_malignant)

# Plot the ROC curve
autoplot(roc_data) +
  labs(
    title = "ROC Curve for Logistic Regression Model",
    subtitle = "Predicting malignant vs. benign tumors"
  )
```



## Plot Manual ROC Curve along with yardstick package calculated ROC

```{r}

# Calculate the automatic yardstick ROC curve for comparison
yardstick_roc_data <- biopsy_predictions %>%
  roc_curve(truth = outcome, .pred_malignant) %>%
  # Use dplyr::select() to ensure the correct function is called
  dplyr::select(-.threshold) %>%
  dplyr::rename(`1-specificity` = specificity) %>%
  dplyr::mutate(`1-specificity` = 1 - `1-specificity`) %>%
  dplyr::mutate(Source = "yardstick")

# Add the manual ROC data to the comparison dataframe
plot_data <- manual_roc_data %>%
  mutate(Source = "Manual Calculation") %>%
  bind_rows(yardstick_roc_data)

# Plot both ROC curves
ggplot(plot_data, aes(x = `1-specificity`, y = sensitivity, color = Source)) +
  geom_line(size = 1) +
  geom_abline(linetype = "dashed") +
  labs(
    title = "ROC Curve: Manual vs. yardstick",
    subtitle = "Comparing calculation methods",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal() +
  theme_minimal()
```

### Kernel Smoothing exercise

```{r}
# Sample data

biopsy_roc_df <- manual_roc_data |> drop_na()
names(biopsy_roc_df) <- c("threshold", "sensitivity", "one_minus_specificity")

threshold <- biopsy_roc_df$threshold
sensitivity <- biopsy_roc_df$sensitivity
one_minus_specificity <- biopsy_roc_df$one_minus_specificity

# Impute or remove NA for smoothing
threshold_clean <- threshold[!is.na(one_minus_specificity)]
one_minus_spec_clean <- one_minus_specificity[!is.na(one_minus_specificity)]

# Kernel smoothing
smooth_1mspec <- ksmooth(x = threshold_clean, y = one_minus_spec_clean,
                         kernel = "normal", bandwidth = 0.02, x.points = threshold)

# Estimate specificity
smooth_spec <- data.frame(
  threshold = smooth_1mspec$x,
  specificity = 1 - smooth_1mspec$y
)

# Optional: smooth sensitivity too
smooth_sens <- ksmooth(x = threshold, y = sensitivity,
                       kernel = "normal", bandwidth = 0.02, x.points = threshold)

# Combine for parameterized curve
roc_curve <- data.frame(
  threshold = threshold,
  sensitivity = smooth_sens$y,
  specificity = 1 - smooth_1mspec$y
)
```



