---
title: "ROC2"
format: html
---


```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "R Packages"
library(tidyverse)
library(tidymodels)
library(MASS)
library(fda)
library(threejs)

```

```{r}
# Load the biopsy dataset
data(biopsy, package = "MASS")

names(biopsy) <- c("ID",
"clump thickness",
"uniformity of cell size",
"uniformity of cell shape",
"marginal adhesion",
"single epithelial cell size",
"bare nuclei",
"bland chromatin",
"normal nucleoli",
"mitoses",
"class")


# Clean column names and convert the outcome to a factor
biopsy <- biopsy %>%
  janitor::clean_names() %>%
  mutate(outcome = factor(class, levels = c("benign", "malignant")))
```


### Fit the Predictive Model

```{r}
# Split the data (a best practice, though not strictly required for this example)
biopsy_split <- initial_split(biopsy, prop = 0.8)
biopsy_train <- training(biopsy_split)
biopsy_test <- testing(biopsy_split)

# Define and fit a logistic regression model
logistic_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(outcome ~ clump_thickness + uniformity_of_cell_size + uniformity_of_cell_shape +
        marginal_adhesion + bare_nuclei, data = biopsy_train)

# Generate predictions on the test set
biopsy_predictions <- predict(logistic_fit, new_data = biopsy_test, type = "prob") %>%
  bind_cols(biopsy_test) %>% drop_na
```

### Manually Create the ROC Curve
```{r}
# Create a sequence of thresholds from 0 to 1
thresholds <- seq(0, 1, by = 0.01)

# Calculate TPR and FPR for each threshold
manual_roc_data <- map_dfr(thresholds, ~ {
  # For a given threshold, classify predictions as "malignant" or "benign"
  predictions_at_threshold <- biopsy_predictions %>%
    mutate(
      .pred_class = factor(
        if_else(.pred_malignant >= .x, "malignant", "benign"),
        levels = c("benign", "malignant")
      )
    )

  # Create a confusion matrix
  conf_mat <- conf_mat(predictions_at_threshold, truth = outcome, estimate = .pred_class)

  # Extract true positives (TP), false positives (FP), etc.
  tp <- conf_mat$table["malignant", "malignant"]
  fp <- conf_mat$table["benign", "malignant"]
  tn <- conf_mat$table["benign", "benign"]
  fn <- conf_mat$table["malignant", "benign"]

  # Calculate TPR and FPR
  tpr <- tp / (tp + fn)
  fpr <- fp / (fp + tn)

  # Return a tibble row for this threshold
  tibble(
    .threshold = .x,
    sensitivity = tpr, # TPR is equivalent to sensitivity
    `1-specificity` = fpr # FPR is equivalent to 1-specificity
  )
})
```

## Plot Manual ROC Curve
```{r}

# Plot the Manual ROC Curve


ggplot(manual_roc_data, aes(x = `1-specificity`, y = sensitivity)) +
  geom_line(size = 1) +
  geom_abline(linetype = "dashed") +
  scale_x_continuous(limits = c(0,0.2)) +
  labs(
    title = "Manual ROC Curve",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal()

```


## Prepare the Data

```{r}
biopsy_roc_df <- manual_roc_data |> drop_na()
names(biopsy_roc_df) <- c("threshold", "sensitivity", "one_minus_specificity")

threshold <- biopsy_roc_df$threshold
sensitivity <- biopsy_roc_df$sensitivity
one_minus_specificity <- biopsy_roc_df$one_minus_specificity

```

### Define a B-spline basis

```{r}
# Choose number of basis functions (adjust for smoothness)
nbasis <- 5
basis <- create.bspline.basis(rangeval = range(threshold), nbasis = nbasis)

```


### Fit FDA Objects

```{r}
# Sensitivity
sens_fd <- smooth.basis(threshold, sensitivity, basis)$fd

# 1 - Specificity
spec_fd <- smooth.basis(threshold, one_minus_specificity, basis)$fd

```

### Evaluate and Visualize

```{r}
# Evaluate at fine grid
grid <- seq(min(threshold), max(threshold), length.out = 100)
sens_vals <- eval.fd(grid, sens_fd)
spec_vals <- eval.fd(grid, spec_fd)

# Plot
plot(grid, sens_vals, type = "l", col = "blue", ylim = c(0,1), ylab = "Value", xlab = "Threshold")
lines(grid, spec_vals, col = "red")
legend("bottomright", legend = c("Sensitivity", "1 - Specificity"), col = c("blue", "red"), lty = 1)

```
### Plot the FDA ROC curve

We now have a differentiable parameterization 
$\gamma(t)=(1-specificity(t), sensitivity(𝑡))$
 suitable for computing arc length, curvature, or tangent vectors.

```{r}
# Plot ROC curve: sensitivity vs. 1-specificity
plot(spec_vals, sens_vals, type = "l", col = "blue", lwd = 2,
     xlab = "1 - Specificity", ylab = "Sensitivity",
     main = "Smoothed ROC Curve (FDA)")
grid()
```

### Diagnostics
```{r}
## Check for NA in grid
anyNA(grid)  # Should return FALSE

#if true rebuild grid explicitly
#grid <- seq(min(threshold, na.rm = TRUE), max(threshold, na.rm = TRUE), length.out = 100)

```

Sometimes smooth.basis() returns an fd object with internal NA values if the input data is sparse or constant (e.g., all zeros for 1-specificity). You can check:

```{r}
summary(spec_fd)
summary(sens_fd)

```
If spec_fd is flat (e.g., all coefficients zero), try increasing nbasis or using a regularization parameter:

```{r}
lambda <- 1e-4
fdPar_obj <- fdPar(basis, Lfdobj = int2Lfd(2), lambda = lambda)
spec_fd <- smooth.basis(threshold_clean, one_minus_spec_clean, fdPar_obj)$fd

```






### Estimating curvature
```{r}
# Assume you already have:
# sens_fd: functional data object for sensitivity
# spec_fd: functional data object for 1 - specificity

# Define evaluation grid
grid <- seq(min(threshold, na.rm = TRUE), max(threshold, na.rm = TRUE), length.out = 100)

# Define derivative operators
D1 <- int2Lfd(1)  # First derivative
D2 <- int2Lfd(2)  # Second derivative

# Evaluate derivatives
x1 <- eval.fd(grid, spec_fd, D1)   # x'(t)
x2 <- eval.fd(grid, spec_fd, D2)   # x''(t)
y1 <- eval.fd(grid, sens_fd, D1)   # y'(t)
y2 <- eval.fd(grid, sens_fd, D2)   # y''(t)

# Curvature formula
curvature <- (x1 * y2 - y1 * x2) / ((x1^2 + y1^2)^(3/2))

# Plot curvature vs. threshold
plot(grid, curvature, type = "l", col = "darkgreen", lwd = 2,
     xlab = "Threshold", ylab = "Curvature",
     main = "Curvature of ROC Curve")
grid()

```

## Combined ROC and Curvature Plot

To plot both the ROC curve and its curvature on the same figure, we’ll use a dual-axis approach in R. This allows you to visualize sensitivity vs. 1-specificity on one axis, and curvature vs. threshold on the other — all aligned by threshold.

```{r}
# Evaluate ROC components
grid <- seq(min(threshold, na.rm = TRUE), max(threshold, na.rm = TRUE), length.out = 100)
sens_vals <- eval.fd(grid, sens_fd)
spec_vals <- eval.fd(grid, spec_fd)

# Derivatives for curvature
D1 <- int2Lfd(1)
D2 <- int2Lfd(2)
x1 <- eval.fd(grid, spec_fd, D1)
x2 <- eval.fd(grid, spec_fd, D2)
y1 <- eval.fd(grid, sens_fd, D1)
y2 <- eval.fd(grid, sens_fd, D2)
curvature <- (x1 * y2 - y1 * x2) / ((x1^2 + y1^2)^(3/2))

# Set up plot window
par(mar = c(5, 4, 4, 4) + 0.1)  # Extra space for right axis

# Plot ROC curve
plot(spec_vals, sens_vals, type = "l", col = "blue", lwd = 2,
     xlab = "1 - Specificity", ylab = "Sensitivity",
     main = "ROC Curve with Curvature Overlay")

# Add curvature on right axis
par(new = TRUE)
plot(grid, curvature, type = "l", col = "darkgreen", lwd = 2, axes = FALSE,
     xlab = "", ylab = "")
axis(side = 4, col = "darkgreen", col.axis = "darkgreen")
mtext("Curvature", side = 4, line = 3, col = "darkgreen")

```




```{r}
# Define curvature threshold for "flat"
epsilon <- 1e-4

# Classify curvature
curv_type <- rep("flat", length(curvature))
curv_type[curvature > epsilon] <- "positive"
curv_type[curvature < -epsilon] <- "negative"

# Plot ROC curve with segmented colors
plot(spec_vals, sens_vals, type = "n",
     xlab = "1 - Specificity", ylab = "Sensitivity",
     main = "ROC Curve Segmented by Curvature Type")

# Segment and plot each type
for (i in 2:length(grid)) {
  seg_color <- switch(curv_type[i],
                      "positive" = "forestgreen",
                      "negative" = "firebrick",
                      "flat"     = "gray50")
  lines(spec_vals[(i-1):i], sens_vals[(i-1):i], col = seg_color, lwd = 2)
}

# Add legend
legend("bottomright", legend = c("Convex", "Concave", "Flat"),
       col = c("forestgreen", "firebrick", "gray50"), lty = 1, lwd = 2)

```

## 3D plot of Curvature Surface


```{r}
grid <- seq(min(threshold, na.rm = TRUE), max(threshold, na.rm = TRUE), length.out = 400)

# Evaluate ROC components
x_vals <- eval.fd(grid, spec_fd)         # 1 - specificity
y_vals <- eval.fd(grid, sens_fd)         # sensitivity

# Derivatives for curvature
D1 <- int2Lfd(1)
D2 <- int2Lfd(2)
x1 <- eval.fd(grid, spec_fd, D1)
x2 <- eval.fd(grid, spec_fd, D2)
y1 <- eval.fd(grid, sens_fd, D1)
y2 <- eval.fd(grid, sens_fd, D2)

# Curvature
z_vals <- (x1 * y2 - y1 * x2) / ((x1^2 + y1^2)^(3/2))

# Combine into a 3-column matrix
roc_matrix <- cbind(x_vals, y_vals, z_vals)

# Normalize curvature for color gradient
curv_scaled <- (z_vals - min(z_vals)) / (max(z_vals) - min(z_vals))
color_map <- rgb(curv_scaled, 1 - curv_scaled, 0.5)

# Plot 3D ROC surface
scatterplot3js(x = roc_matrix,
               color = color_map,
               size = 0.1,
               labels = round(grid, 3),
               xlab = "1 - Specificity",
               ylab = "Sensitivity",
               zlab = "Curvature")


```







Clinical Tip
If 1-specificity is flat or sparse, consider modeling it jointly with sensitivity using a multivariate FDA approach or regularized smoothing. I can help you build a joint basis or penalized functional regression if needed.



