---
title: "ROC3"
date: September 29, 2025
format: html
---


```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "R Packages"
library(tidyverse)
library(tidymodels)
library(MASS)
library(fda)
library(threejs)

```

```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the Code"

# Load the biopsy dataset
data(biopsy, package = "MASS")

names(biopsy) <- c("ID",
"clump thickness",
"uniformity of cell size",
"uniformity of cell shape",
"marginal adhesion",
"single epithelial cell size",
"bare nuclei",
"bland chromatin",
"normal nucleoli",
"mitoses",
"class")


# Clean column names and convert the outcome to a factor
biopsy <- biopsy %>%
  janitor::clean_names() %>%
  mutate(outcome = factor(class, levels = c("malignant", "benign")))
```


### Fit the Predictive Model

```{r}
#| code-fold: TRUE
#| code-summary: "Show the Code"
# Split the data (a best practice, though not strictly required for this example)
biopsy_split <- initial_split(biopsy, prop = 0.8)
biopsy_train <- training(biopsy_split)
biopsy_test <- testing(biopsy_split)

# Define and fit a logistic regression model
logistic_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(outcome ~ clump_thickness + uniformity_of_cell_size + uniformity_of_cell_shape +
        marginal_adhesion + bare_nuclei, data = biopsy_train)

# Generate predictions on the test set
biopsy_predictions <- predict(logistic_fit, new_data = biopsy_test, type = "prob") %>%
  bind_cols(biopsy_test) %>% drop_na()
```

### Manually Create the ROC Curve
```{r}
#| code-fold: TRUE
#| code-summary: "Show the Code"
# Create a sequence of thresholds from 0 to 1
thresholds <- seq(0, 1, by = 0.01)

# Calculate TPR and FPR for each threshold
manual_roc_data <- map_dfr(thresholds, ~ {
  # For a given threshold, classify predictions as "malignant" or "benign"
  predictions_at_threshold <- biopsy_predictions %>%
    mutate(
      .pred_class = factor(
        if_else(.pred_malignant >= .x, "malignant", "benign"),
        levels = c("malignant", "benign")
      )
    )

  # Create a confusion matrix
  conf_mat <- conf_mat(predictions_at_threshold, truth = outcome, estimate = .pred_class)

  # Extract true positives (TP), false positives (FP), etc.
  tp <- conf_mat$table["malignant", "malignant"]
  fp <- conf_mat$table["benign", "malignant"]
  tn <- conf_mat$table["benign", "benign"]
  fn <- conf_mat$table["malignant", "benign"]

  # Calculate TPR and FPR
  tpr <- tp / (tp + fn)
  fpr <- fp / (fp + tn)

  # Return a tibble row for this threshold
  tibble(
    .threshold = .x,
    sensitivity = tpr, # TPR is equivalent to sensitivity
    `1-specificity` = fpr # FPR is equivalent to 1-specificity
  )
})

manual_roc_data <- manual_roc_data %>% drop_na()
```

## Plot Manual ROC Curve
```{r}
#| code-fold: TRUE
#| code-summary: "Show the Code"
# Plot the Manual ROC Curve


ggplot(manual_roc_data, aes(x = `1-specificity`, y = sensitivity)) +
  geom_line(size = 1) +
  geom_abline(linetype = "dashed") +
  scale_x_continuous(limits = c(0,0.2)) +
  labs(
    title = "Manual ROC Curve",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal()

```
### Tidyverse ROC Calculation
```{r}
#| code-fold: TRUE
#| code-summary: "Show the Code"
# Calculate ROC curve data
roc_data <- biopsy_predictions %>%
  roc_curve(truth = outcome, .pred_malignant)

# Plot the ROC curve
autoplot(roc_data) +
  labs(
    title = "ROC Curve for Logistic Regression Model",
    subtitle = "Predicting malignant vs. benign tumors"
  )
```



## Prepare the Data for FDA ROC

```{r}
#| code-fold: TRUE
#| code-summary: "Show the Code"
biopsy_roc_df <- manual_roc_data |> drop_na()
names(biopsy_roc_df) <- c("threshold", "sensitivity", "one_minus_specificity")

threshold <- biopsy_roc_df$threshold
sensitivity <- biopsy_roc_df$sensitivity
one_minus_specificity <- biopsy_roc_df$one_minus_specificity

```

### Tidyverse Code to Prepare Data for FDA

```{r}
#| code-fold: TRUE
#| code-summary: "Show the Code"

# roc_data_2 <- roc_data %>% filter(row_number() != 1 & row_number() != n())
# 
# roc_data_3 <- roc_data_2 %>% mutate(one_minus_specificity = 1- specificity)
# names(roc_data_3) <- c("threshold", "specificity", "sensitivity", "one_minus_specificity")
# 
# 
# threshold <- roc_data_3$threshold
# sensitivity <- roc_data_3$sensitivity
# one_minus_specificity <- roc_data_3$one_minus_specificity

```







```{r}
 #Create B-spline basis
#| code-fold: TRUE
#| code-summary: "Show the Code"
nbasis <- 5
basis <- create.bspline.basis(rangeval = range(threshold), nbasis = nbasis)

# Fit FDA objects
sens_fd <- smooth.basis(threshold, sensitivity, basis)$fd
spec_fd <- smooth.basis(threshold, one_minus_specificity, basis)$fd

# Evaluate on fine grid
grid <- seq(min(threshold), max(threshold), length.out = 100)
sens_vals <- eval.fd(grid, sens_fd)
spec_vals <- eval.fd(grid, spec_fd)

# Plot ROC curve: sensitivity vs. 1-specificity
plot(spec_vals, sens_vals, type = "l", col = "blue", lwd = 2,
     xlab = "1 - Specificity", ylab = "Sensitivity",
     main = "Smoothed ROC Curve (FDA)")
grid()
```

### Estimate Curvature
```{r}
#| code-fold: TRUE
#| code-summary: "Show the Code"
library(fda)

# Clean threshold grid
grid <- seq(min(threshold, na.rm = TRUE), max(threshold, na.rm = TRUE), length.out = 100)

# Derivative operators
D1 <- int2Lfd(1)
D2 <- int2Lfd(2)

# Evaluate derivatives
x1 <- eval.fd(grid, spec_fd, D1)
x2 <- eval.fd(grid, spec_fd, D2)
y1 <- eval.fd(grid, sens_fd, D1)
y2 <- eval.fd(grid, sens_fd, D2)

# Curvature formula
curvature <- (x1 * y2 - y1 * x2) / ((x1^2 + y1^2)^(3/2))

# Plot
plot(grid, curvature, type = "l", col = "darkgreen", lwd = 2,
     xlab = "Threshold", ylab = "Curvature",
     main = "Curvature of ROC Curve")
grid()


```

```{r}
#| code-fold: TRUE
#| code-summary: "Show the Code"
curvature_df <- data.frame(grid, x1, x2, y1, y2, curvature)
```





`




## 3D plot of Curvature Surface


```{r}
grid <- seq(min(threshold, na.rm = TRUE), max(threshold, na.rm = TRUE), length.out = 400)

# Evaluate ROC components
x_vals <- eval.fd(grid, spec_fd)         # 1 - specificity
y_vals <- eval.fd(grid, sens_fd)         # sensitivity

# Derivatives for curvature
D1 <- int2Lfd(1)
D2 <- int2Lfd(2)
x1 <- eval.fd(grid, spec_fd, D1)
x2 <- eval.fd(grid, spec_fd, D2)
y1 <- eval.fd(grid, sens_fd, D1)
y2 <- eval.fd(grid, sens_fd, D2)

# Curvature
z_vals <- (x1 * y2 - y1 * x2) / ((x1^2 + y1^2)^(3/2))

# Combine into a 3-column matrix
roc_matrix <- cbind(x_vals, y_vals, z_vals)

# Normalize curvature for color gradient
curv_scaled <- (z_vals - min(z_vals)) / (max(z_vals) - min(z_vals))
color_map <- rgb(curv_scaled, 1 - curv_scaled, 0.5)

# Plot 3D ROC surface
scatterplot3js(x = roc_matrix,
               color = color_map,
               size = 0.1,
               labels = round(grid, 3),
               xlab = "1 - Specificity",
               ylab = "Sensitivity",
               zlab = "Curvature")


```







Clinical Tip
If 1-specificity is flat or sparse, consider modeling it jointly with sensitivity using a multivariate FDA approach or regularized smoothing. I can help you build a joint basis or penalized functional regression if needed.

### Diagnostics
```{r}
#| code-fold: TRUE
#| code-summary: "R Packages"
## Check for NA in grid
anyNA(grid)  # Should return FALSE

#if true rebuild grid explicitly
#grid <- seq(min(threshold, na.rm = TRUE), max(threshold, na.rm = TRUE), length.out = 100)

```

Sometimes smooth.basis() returns an fd object with internal NA values if the input data is sparse or constant (e.g., all zeros for 1-specificity). You can check:

```{r}
#| code-fold: TRUE
#| code-summary: "R Packages"
summary(spec_fd)
summary(sens_fd)

```
If spec_fd is flat (e.g., all coefficients zero), try increasing nbasis or using a regularization parameter:

```{r}
#| code-fold: TRUE
#| code-summary: "R Packages"
lambda <- 1e-4
fdPar_obj <- fdPar(basis, Lfdobj = int2Lfd(2), lambda = lambda)
spec_fd <- smooth.basis(threshold, one_minus_specificity, fdPar_obj)$fd

```

