---
title: "asthma4state"
format: html
---

This workbook attempts to find a symbolic solution to the  Kolmogorov Forward equation given on page 346 of [Network Analysis for Decision Making](https://www.agropustaka.id/wp-content/uploads/2020/04/agropustaka.id_buku_Network-Meta-Analysis-for-Decision-Making.pdf) by Dias et al. 

THe strategy is to use the `reticulate` R package to access the symbolic engine in the python `caracas` package.

## Setup

* Install the `reticulate` package
* Install the `caracas` R package
* Run the command: `caracas::install_sympy()`

If it doesn't work run:  `reticulate::py_config()` to see your python configuration

If for some reason it does not work, try updating conda with this R command: `reticulate::miniconda_update()`



## 4 state Asthma Model


$$ G = 
	\begin{bmatrix} 
	-\lambda_{12} & \lambda_{12} & 0 & 0 \\
	\lambda_{21} & -(\lambda_{21} + \lambda_{23} + \lambda_{24}) & \lambda_{23} & \lambda_{24}\\
	 0 & \lambda_{32} & -(\lambda_{32} + \lambda_{34}) & \lambda_{34} \\
	 0 & 0 & 0 & 1\\
	\end{bmatrix}
	\quad
	$$

$$ \frac{dP(u)}{du} = P(u)G$$


```{r}
#| message: FALSE
#| warning: False
#| echo: FALSE
library(caracas)
caracas::install_sympy()
library(tidyverse)
library(reticulate)
```



## Failed attempt to use the caracas R package

```{r}
#| message: FALSE
#| warning: False
#| include: FALSE
G <- matrix(c("-lam12", "lam12", "0", "0",
                "lam21", "-(lam21 + lam23 + lam24)", "lam23", "lam24", 
                  "0", "lam32", "-(lam32 + lam34)" , "lam34",
                  "0", "0", "0", "0"), 4, 4, byrow=TRUE) %>% as_sym()
G
u <- symbol("u")
```

### Tried a whole bunch of stuff that didn't work
```{r}
#| message: FALSE
#| warning: False
#| include: FALSE
 # Compute the matrix exponential e^{uG}
 expG <- expm(u * G)

#Error in expm(u * G) : is.numeric(x) || is.complex(x) || is.logical(x) || (isM <- inherits(x,  .... is not TRUE

 # Compute the matrix exponential e^{uG}
 expG <- caracas::expm(u * G)
#Error: 'expm' is not an exported object from 'namespace:caracas'
 
 expG <- G$exp()  # This computes the symbolic matrix exponential of G
#Error: attempt to apply non-function
 
# Compute the matrix exponential e^{uG}
# Multiply the matrix by u symbolically
 Gu <- u * G
 expGu <- Gu$exp()
#Error: attempt to apply non-function
 
 # Compute the matrix exponential e^{uG}
 # Multiply the matrix by u symbolically
 Gu <- u * G
expGu <- caracas:::sympy_function_call(Gu, "exp")
#Error: object 'sympy_function_call' not found

G <- matrix(c("-lam12", "lam12", "0", "0",
                "lam21", "-(lam21 + lam23 + lam24)", "lam23", "lam24", 
                 "0", "lam32", "-(lam32 + lam34)" , "lam34"), 3, 4, byrow=TRUE) %>% as_sym()
G

u <- symbol("u")
Gu

# Compute the matrix exponential e^{uG}
# Multiply the matrix by u symbolically
Gu <- u * G
expGu <- caracas:::sympy_function_call(Gu, "exp")
Gu <- u * G
 
# Convert to character so we can pass it to SymPy
 Gu_str <- as.character(Gu)
 
 # Compute the matrix exponential using SymPy directly
expGu <- call_sympy(paste0("(", Gu_str, ").exp()"))
#Error in call_sympy(paste0("(", Gu_str, ").exp()")) : could not find function "call_sympy"

```


## Attempt to use `reticulate` to call Python.
This ran for 45 minutes on my Macbook Air but did not terminate.
(1.1GHz, quad-core, Intel chip, 16MB memory, macOS Sequoia 15.5)

```{python}
from sympy import symbols, Matrix, exp
u, λ12, λ21, λ23, λ24, λ32, λ34 = symbols('u lambda12 lambda21 lambda23 lambda24 lambda32 lambda34')
G = Matrix([
    [-λ12,       λ12,             0,        0],
    [λ21, - (λ21 + λ23 + λ24),  λ23,      λ24],
    [0,           λ32, - (λ32 + λ34),     λ34],
    [0,             0,            0,        0]
])
Gu = u * G
expGu = Gu.exp()
expGu

```


### Environment check
```{python}
# Use a simpler test matrix to make sure the pipeline works
from sympy import symbols, Matrix
u, λ = symbols('u lambda')
G_test = Matrix([[ -λ, λ], [λ, -λ]])
Gu_test = u * G_test
Gu_test_exp = Gu_test.exp()
Gu_test_exp

```


### Test feasibility before running

Bing AI advises: If your matrix is larger than 3×3 and filled with symbols, symbolic exponentiation is likely to be intractable.

Try `G.is_diagonalizable()` in SymPy: If it returns True, you can use diagonalization to compute the exponential more efficiently. Tollowing code did not terminate

```{python}
G.is_diagonalizable()
```

Use `.eigenvals()` or `.jordan_form()`: If these return quickly, symbolic exponentiation might be feasible via spectral methods.

```{python}
G.eigenvals(G)
```
This result indicates distinct eigenvalues with respective multiplicity 1, 1, 1, 1. These eigenvalues control the rates of convergence to the stationary distribution. The slowest decay rate is governed by the eigenvalue closest to zero (-0.171), which influences the long-term behavior.

This code did not terminate
```{python}
G.jordan_form()
```



### Hybrid Method
Substitute numeric values for some or all $\lambda_{ij}$ to reduce symbolic complexity.
G is random Q matirx. u is still symbolic.

The following code did not terminate.

```{python}
from sympy import symbols, Matrix
u = symbols('u')
G = Matrix([
    [-0.5, 0.5, 0, 0],
    [1.0, -2.0, 0.6, 0.4],
    [0, 1.2, -2.5, 1.3],
    [0, 0, 0, 0]
])
Gu = u * G
expGu = Gu.exp()
expGu

```


### Numerical example with random Q matrix, G

```{r}
library(expm)
library(ggplot2)
library(tidyr)
library(dplyr)

# Define G numerically
G <- matrix(c(
  -0.5, 0.5, 0, 0,
   1.0, -2.0, 0.6, 0.4,
   0, 1.2, -2.5, 1.3,
   0, 0, 0, 0
), nrow = 4, byrow = TRUE)

# Initial state
P0 <- c(1, 0, 0, 0)

# Time points
u_vals <- seq(0, 10, by = 0.1)

# Compute P(u) over time
P_matrix <- sapply(u_vals, function(u) P0 %*% expm(u * G))

# Format for plotting
df <- as.data.frame(t(P_matrix))
colnames(df) <- paste0("State_", 1:4)
df$Time <- u_vals
df_long <- pivot_longer(df, cols = starts_with("State"), names_to = "State", values_to = "Probability")

# Plot
ggplot(df_long, aes(x = Time, y = Probability, color = State)) +
  geom_line(size = 1.2) +
  labs(title = "Evolution of State Probabilities Over Time", y = "Probability", x = "Time") +
  theme_minimal()

```
### Symbolic Eigenvectors in Python (SymPy)

This runs but produces unreadable output
```{python}
from sympy import symbols, Matrix

# Define symbols
λ12, λ21, λ23, λ24, λ32, λ34 = symbols('lambda12 lambda21 lambda23 lambda24 lambda32 lambda34')

# Define G
G = Matrix([
    [-λ12, λ12, 0, 0],
    [λ21, -(λ21 + λ23 + λ24), λ23, λ24],
    [0, λ32, -(λ32 + λ34), λ34],
    [0, 0, 0, 0]
])

# Compute eigenvalues and eigenvectors
eigen_data = G.eigenvects()

# Display eigenvalues and corresponding eigenvectors
for val, mult, vects in eigen_data:
    print(f"Eigenvalue: {val}")
    for v in vects:
        print(f"Eigenvector: {v}")

```


```{python}
P, J = G.jordan_form()
expGu = P * (u * J).exp() * P.inv()

```


