---
title: "new_3state_solution"
format: html
---

This document implements a new approach to solving the Kolmogorov Forward Equations for the 3-state absorbing continuous time Markov chain given an aggregate state table computed a a particular time. The approach is as follows:

1. Implement a JAGS model to estimate the generator matrix 𝑄 from the state table.
2. Solve the differential equations for the Kolmogorov Forward Equations using the estimated parameters from the JAGS model.


### Load the required libraries

```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"
library(rjags)
library(ggplot2)
library(coda)
library(dplyr)
library(tidyr)
```

### Input the data

```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"
# Example state table: ST[i,j] = transitions from i to j
ST <- matrix(c(210,  61,  1,
                89, 645, 14,
                0,    0, 81),  nrow=3, ncol=3, byrow=TRUE)

# ST <- matrix(c(
#   10, 12, 5,   # from state 1: 12 to state 2, 5 to state 3
#   8, 10, 7,    # from state 2: 8 to state 1, 7 to state 3
#   0, 0, 20    # from state 3: 20 self-transitions, no exits
# ), nrow = 3, byrow = TRUE)

# Total time spent in states 1 and 2
T_vec <- rowSums(ST)[1:2]  # c(17, 15)

data_jags <- list(
  ST = ST,
  T = as.numeric(T_vec)  # Ensure it's a numeric vector
)


```

### The JAGS Model

```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"

model_string <- "
model {
  # Likelihood: Poisson transitions from state i to j
  for (i in 1:2) {
    for (j in 1:3) {
      ST[i,j] ~ dpois(rate[i,j])
    }
  }

  # State 3 is absorbing: no transitions out
  for (j in 1:3) {
    rate[3,j] <- 0
  }

  # Define rates from Q matrix
  rate[1,1] <- 0
  rate[1,2] <- q12 * T[1]
  rate[1,3] <- q13 * T[1]

  rate[2,1] <- q21 * T[2]
  rate[2,2] <- 0
  rate[2,3] <- q23 * T[2]

  # Priors for off-diagonal rates
  q12 ~ dgamma(1, 1)
  q13 ~ dgamma(1, 1)
  q21 ~ dgamma(1, 1)
  q23 ~ dgamma(1, 1)

  # Construct generator matrix Q
  Q[1,1] <- - (q12 + q13)
  Q[1,2] <- q12
  Q[1,3] <- q13

  Q[2,1] <- q21
  Q[2,2] <- - (q21 + q23)
  Q[2,3] <- q23

  Q[3,1] <- 0
  Q[3,2] <- 0
  Q[3,3] <- 0
}


"


writeLines(model_string, "new_3state_solution.bug")
```

### Run the JAGS Model

```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"
# Initialize model
# Compile model
jags_model <- jags.model(file = "new_3state_solution.bug",
                         data = data_jags,
                         n.chains = 3,
                         n.adapt = 1000)

# Burn-in
update(jags_model, 1000)
```

Extract Samples
```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"
# Sample from posterior
samples <- coda.samples(jags_model,
                        variable.names = c("Q"),
                        n.iter = 5000)

```

### Summarize the Samples
```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"
samples_df <- as.data.frame(as.mcmc(do.call(rbind, samples)))
smry <- summary(samples)
Mean <- smry$statistics[, "Mean"]
Q_smry <- cbind(Mean, smry$quantiles)
round(Q_smry, 4)
```





## Set up fpr MCMC Trace Plots

```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"
# Convert mcmc.list to data frame and add chain + iteration info
df_long <- do.call(rbind, lapply(1:length(samples), function(chain) {
  as.data.frame(samples[[chain]]) %>%
    mutate(
      iteration = row_number(),
      chain = factor(chain)
    )
})) %>%
  pivot_longer(cols = -c(iteration, chain), 
               names_to = "parameter", 
               values_to = "value")
```



### Trace Plots

```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"
# Define the number of rows to sample
n_sample <- 10000

# Take a random sample of n_sample rows
#f) gets the total number of rows in the data frame
# sample(nrow(df), n_sample) generates n_sample random row indices
# df[...] then subsets the data frame using these random indices
random_sample_df <- df_long[sample(nrow(df_long), n_sample), ]
ggplot(random_sample_df %>% filter(parameter %in% c("Q[1,1]", "Q[1,2]", "Q[1,3]", 
                                           "Q[2,1]", "Q[2,2]", "Q[2,3]")), 
       aes(x = iteration, y = value, color = chain)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~parameter, scales = "free_y") +
  labs(title = "Trace Plots", x = "Iterations", y = "Parameter Value") +
  theme_minimal() +
  scale_x_continuous(breaks = n_sample)
```







### Plot Posterior distributions.
```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: true
#| code-summary: "Show the code"
ggplot(df_long %>% filter(parameter %in% c("Q[1,1]", "Q[1,2]", "Q[1,3]", 
                                           "Q[2,1]", "Q[2,2]", "Q[2,3]")), 
       aes(x = value, fill = parameter)) +
       geom_density(alpha = 0.5) +
       facet_wrap(~parameter, scales = "free") +
       xlim(0,1) +
       labs(title = "Posterior Densities", x = "Value", y = "Density") +
       theme_minimal()
```


Functions to compute transition Probabilities

```{r}
# Define symbolic transition probability functions
compute_P <- function(t, q12, q13, q21, q23) {
  # Define shorthand
  alpha <- q12 + q13
  beta  <- q21 + q23
  delta <- sqrt((alpha - beta)^2 + 4 * q12 * q21)

  lambda1 <- (- (alpha + beta) + delta) / 2
  lambda2 <- (- (alpha + beta) - delta) / 2

  # Transition probabilities from state 1
  P11 <- ((lambda1 + alpha) / (lambda1 - lambda2)) * exp(lambda2 * t) +
         ((lambda2 + alpha) / (lambda2 - lambda1)) * exp(lambda1 * t)

  P12 <- (q12 / (lambda1 - lambda2)) * (exp(lambda1 * t) - exp(lambda2 * t))
  P13 <- 1 - P11 - P12

  # Transition probabilities from state 2
  P21 <- (q21 / (lambda1 - lambda2)) * (exp(lambda1 * t) - exp(lambda2 * t))

  P22 <- ((lambda1 + beta) / (lambda1 - lambda2)) * exp(lambda2 * t) +
         ((lambda2 + beta) / (lambda2 - lambda1)) * exp(lambda1 * t)

  P23 <- 1 - P21 - P22

  # Transition probabilities from absorbing state 3
  P31 <- 0
  P32 <- 0
  P33 <- 1

  # Assemble full transition matrix
  P <- matrix(c(
    P11, P12, P13,
    P21, P22, P23,
    P31, P32, P33
  ), nrow = 3, byrow = TRUE)

  return(P)
}

```

Compute Transition Probabilities


```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"
# Define rates
q12 <- mean(samples_df$"Q[1,2]")
q13 <- mean(samples_df$"Q[1,3]")
q21 <- mean(samples_df$"Q[2,1]")
q23 <- mean(samples_df$"Q[2,3]")
t <- 12

P <- compute_P(t, q12, q13, q21, q23)
colnames(P) <- c("State 1","State 2", "State 3")
rownames(P) <- c("State 1", "State 2", "State 3")
round(P,4)
```

## The Differential Equations
This section shows the steps to solve the differential equations for solving the Kolmogorov Forward Equations for the 3-state continuous time markov chain where the third state is an absorbing state.

### Step 1: Define the Generator Matrix  𝑄

We assume:

* States 1 and 2 are transient.
* State 3 is absorbing.
* Generator matrix 𝑄 has the form:

$$Q = 
\begin{bmatrix}
- (q_{12} + q_{13}) & q_{12} & q_{13} \\
q_{21} & - (q_{21} + q_{23}) & q_{23} \\
0 & 0 & 0
\end{bmatrix}$$

$$\alpha = q_{12} + q_{13}, \quad \beta = q_{21} + q_{23}$$



### Step 2: Kolmogorov Forward Equations
$$\frac{d}{dt} \mathbf{p}(t) = Q \mathbf{p}(t), \quad \mathbf{p}(0) = 
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}$$

This gives the system:

$$\begin{aligned}
\frac{dp_1}{dt} &= -\alpha p_1 + q_{21} p_2 \\
\frac{dp_2}{dt} &= q_{12} p_1 - \beta p_2 \\
\frac{dp_3}{dt} &= q_{13} p_1 + q_{23} p_2
\end{aligned}$$



### Step 3: Solve the Coupled System for 𝑝1(𝑡) and 𝑝2(𝑡)
This is a linear system of two first-order ODEs. We write it in matrix form:


$$\frac{d}{dt}
\begin{bmatrix}
p_1(t) \\
p_2(t)
\end{bmatrix}
=
\begin{bmatrix}
- \alpha & q_{21} \\
q_{12} & - \beta
\end{bmatrix}
\begin{bmatrix}
p_1(t) \\
p_2(t)
\end{bmatrix}$$


Let this matrix be 𝐴. The solution is:

$$\begin{bmatrix}
p_1(t) \\
p_2(t)
\end{bmatrix}
= e^{A t}
\begin{bmatrix}
1 \\
0
\end{bmatrix}$$


### Step 4: Diagonalize 𝐴 The eigenvalues of 𝐴are:

$$\lambda_{1,2} = \frac{- (\alpha + \beta) \pm \sqrt{(\alpha - \beta)^2 + 4 q_{12} q_{21}}}{2}$$
Let:

$$\delta = \sqrt{(\alpha - \beta)^2 + 4 q_{12} q_{21}}, \quad
\lambda_1 = \frac{- (\alpha + \beta) + \delta}{2}, \quad
\lambda_2 = \frac{- (\alpha + \beta) - \delta}{2}$$

### Step 5: Solve for 𝑝1(𝑡) and 𝑝2(𝑡)
Using variation of parameters or Laplace transforms, we obtain:

$$\begin{aligned}
p_1(t) &= \left( \frac{\lambda_1 + \alpha}{\lambda_1 - \lambda_2} \right) e^{\lambda_2 t}
+ \left( \frac{\lambda_2 + \alpha}{\lambda_2 - \lambda_1} \right) e^{\lambda_1 t} \\
p_2(t) &= \frac{q_{12}}{\lambda_1 - \lambda_2} \left( e^{\lambda_1 t} - e^{\lambda_2 t} \right)
\end{aligned}$$

### Step 6: Solve for 𝑝3(𝑡) Integrate the third equation:

$$\frac{dp_3}{dt} = q_{13} p_1(t) + q_{23} p_2(t)$$
Then:

$$p_3(t) = \int_0^t \left( q_{13} p_1(s) + q_{23} p_2(s) \right) ds$$

But since total probability must sum to 1:

$$p_3(t) = 1 - p_1(t) - p_2(t)$$

### Final Transition Probabilities from State 1

$$\begin{aligned}
P_{11}(t) &= p_1(t) \\
P_{12}(t) &= p_2(t) \\
P_{13}(t) &= 1 - p_1(t) - p_2(t)
\end{aligned}$$


### Transition Probabilities from State 2
Repeat the same process with initial condition 𝑝(0)=[0,1,0]⊤. The structure is symmetric, and the solutions are:

$$\begin{aligned}
P_{21}(t) &= \frac{q_{21}}{\lambda_1 - \lambda_2} \left( e^{\lambda_1 t} - e^{\lambda_2 t} \right) \\
P_{22}(t) &= \left( \frac{\lambda_1 + \beta}{\lambda_1 - \lambda_2} \right) e^{\lambda_2 t}
+ \left( \frac{\lambda_2 + \beta}{\lambda_2 - \lambda_1} \right) e^{\lambda_1 t} \\
P_{23}(t) &= 1 - P_{21}(t) - P_{22}(t)
\end{aligned}$$



### Transition Probabilities from State 3 (Absorbing)

$$P_{31}(t) = 0, \quad P_{32}(t) = 0, \quad P_{33}(t) = 1$$
